Notes on conformal manifolds

### 2025-06-30 Draft_MLFlatDirections.tex
Update the draft

\title Machine learning to study marginal deformation of 2d conformal field theories

@@

From this perspective, the AdS/CFT correspondence relates conformal manifolds on the boundary theory to flat directions in the scalar potential of the dual supergravity theory. Along these directions, the scalar field configurations vary continuously while the cosmological constant remains fixed. In general, supersymmetry is believed to be necessary for the existence of holographic conformal manifolds, as non-supersymmetric AdS solutions are typically considered unstable \cite{Ooguri:2016pdq,Palti:2019pca}. However, recent investigations \cite{Giambrone:2021wsm} have identified AdS$_4$ configurations that appear to evade this requirement, with no evidence of standard decay channels—neither perturbative nor non-perturbative—being present. In the context of AdS$_3$/CFT$_2$, the situation may be even richer. A well-known counterexample has existed for some time \cite{Aharony:2001dp,Dong:2014tsa}, which is well understood from both the field theory and gravitational perspectives. More generally, within the supergravity approximation, continuous deformations correspond to classically marginal operators. These operators become exactly marginal only if their marginality persists at finite values of both the rank N and the CFT coupling. Recent work \cite{Eloy:2024lwn} has demonstrated the existence of families of marginal deformations that are perturbatively stable, using the framework of Exceptional Field Theory (ExFT).

In this work, we propose a novel approach, where we study directly the scalar potential of supergravity. As already mentionned, the marginal deformations of CFT's in the large N limit, are in correspondance with the flat directions in the potential, i.e. when there is a continuous set a scalar fields such that $\nabla V = 0$. However, the explicit characterization of these flat directions presents formidable technical challenges. Supergravity scalar potentials, even in truncated models, typically involve dozens of scalar fields with intricate non-linear interactions. The resulting expressions for critical points—where all first derivatives vanish—quickly become too complex for traditional symbolic manipulation, rendering analytical approaches computationally intractable. This complexity paradox is particularly frustrating given that physically meaningful solutions often admit relatively simple parametric descriptions once discovered.

The emergence of machine learning techniques in theoretical physics opens new avenues for addressing such complex problems. Instead of solving the full symbolic system analytically from the outset, one can employ numerical methods to sample the solution space and subsequently apply symbolic regression techniques to extract analytical patterns from the data. This hybrid approach has the potential to bypass the computational bottlenecks inherent in purely symbolic methods, while still aiming to uncover exact analytical expressions.

Similar strategies have previously been applied to identify new isolated vacua in $\mathrm{SO}(8)$ supergravity \cite{Comsa:2019rcz,Berman:2022jqn}. More broadly, there has been increasing interest in applying machine learning and numerical techniques across various domains of high-energy physics. This includes, for instance, the characterization of Calabi–Yau metrics and hypersurfaces \cite{Ashmore:2019wzb,Berman:2021mcw,Larfors:2022nep,Berglund:2022gvm,Jejjala:2020wcc,Douglas:2006rr,Larfors:2021pbb,He:2018jtw}, as well as broader efforts to explore the string theory landscape using machine learning techniques \cite{He:2017aed,Carifio:2017bov,Ruehle:2017mzq}. Additional applications include studies in CFT \cite{Chen:2020dxg}, investigations of the supergravity landscape \cite{Brady:2025zzi,Krishnan:2020sfg}, and explorations of the AdS/CFT correspondence \cite{Hashimoto:2018ftp}. More generally, machine learning has found utility in the study of string theory, geometry, and fundamental physics \cite{Ruehle:2020jrk,He:2023csq,Bao:2021auj}.

In this work, we demonstrate the viability of the aformentionned machine learning approach by applying it to a five-scalar subsector of a 13 scalar consistent truncation of three-dimensional chiral $\mathcal{N} = (1,1)$  supergravity arising from $AdS_3 \times S^3$ compactification from 6d supergravity, or from the compactification of type IIB supergravity on $AdS_3 \times S^3 \times T^4$. Our methodology combines gradient descent sampling of the flat direction manifold with some symbolic regression technique. There exist a large litterature on symbolic regression, using methods from genetic programming \cite{koza1994genetic} such as \cite{virgolin2021improving,randall2022bingo,burlacu2019parsimony}, to deep machine learning \cite{petersen2019deep,kamienny2022end}, generative modelling \cite{valipour2021symbolicgpt}, diffusion models \cite{bastiani2025diffusion}, and equation learner with the nodes being symbolic operations \cite{2018arXiv180607259S}. Another state-of-the art algorithm is the AIFeynman methods \cite{Udrescu:2019mnk}. They use neural network to identify structure in the dataset (such as translational symmetries, multiplicative separability, compositionality, \dots) to recursively define simpler problems on which they can fit the solutions with polynomials.  However, in the spirit of trying to build a generalisable method, we develop here our own algorithm, which is based on symbolic regression using an Annealing Importance Sampling method (AIS) \cite{neal1998annealedimportancesampling}. In this paper, we will study a 5 parameter truncation, which is itself coming from a 13 parameter consistent truncation of $\mathcal{N} = (1,1)$ 6-dimensional supergravity.

The study of this 5 dimensional parameter potentiel is separated into several parts. We first use the gradient descent to efficiently sample the underlying conformal manifold. Combined with numerical analysis, such as principal component analysis and clustering, we can identify the existence of a 3-dimensional continuous family. We then use an AIS technique combined with a sequential Monte Carlo approach (SMC) \cite{del2006sequential} in order to do symbolic regression. As we will indeed demonstrate in the paper, there exist polynomial constraints on the 5 parameters, viewed as embedding coordinates, that projects them on the 3 dimensional manifolds. We manage to identify 8 of those constraints on the data, not all independent, which once solved gives that two of the embedding coordinates can be expressed in terms of the others. This provides an explicit family of 3 dimensional parameters family

The paper is organized as follows. Section 2 establishes the supergravity setup, presenting the scalar potential in its full complexity and motivating the truncation to five fields. Section 3 details our numerical methodology, covering both the gradient descent sampling procedure and the annealed importance sampling approach to symbolic regression. Section 4 analyzes the discovered supergravity solutions, including their geometric interpretation and spectrum analysis. Finally, Section 5 concludes with prospects for extending this approach to higher-dimensional cases and its broader implications for systematic studies of conformal manifolds in holographic theories.

@@

The Chern-Simons contribution is given by
%
\begin{equation}
	{\cal L}_{\rm CS}= -\varepsilon^{\,\mu\nu\rho}\,\Theta_{\bar M\bar N|\bar P\bar Q}\,A_{\mu}{}^{\bar M\bar N}\left(\partial_{\nu}\,A_{\rho}{}^{\bar P\bar Q}  + \frac{1}{3}\, \Theta_{\bar R\bar S|\bar U\bar V}\,f^{\bar P\bar Q,\bar R\bar S}{}_{\bar X\bar Y}\, A_{\nu}{}^{\bar U\bar V} A_{\rho}{}^{\bar X\bar Y} \right)\,,
\end{equation}
%
where $f^{\bar M\bar N,\bar P\bar Q}{}_{\bar K\bar L} = 4\,\delta_{[\bar K}{}^{[\bar M}\eta^{\bar N][\bar P}\delta_{\bar L]}{}^{\bar Q]}$ are the structure constants of the SO(8,4) Lie algebra, and $\varepsilon^{\mu\nu\rho}$ is the three-dimensional Levi-Civita tensor.

@@

As an initial simplification, we restrict our analysis to the scalar fields $x_{1}, x_{2}, x_{4}, \tilde{x}_{8}$, and $x_{10}$. This subset, selected at random from the original set of thirteen scalar fields, serves to render the problem more tractable at the outset. It is important to emphasize, however, that this subset does not constitute a consistent truncation of the full theory. Consequently, the effective theory defined by these five scalars may be formally ill-defined, as higher-order interactions could, in principle, source the remaining fields.

@@

		To do so, we perform a basic gradient descent. We begin by randomly and uniformly initializing points within a hypercube of range $[-2,2]$. This choice is important to ensure that points are not restricted to the inner range $[-1,1]$. This will be important later, when performing the symbolic regression. For this example, we choose to generate $10^5$ points. This number is motivated by the dimensionality of our problem. If we aim to populate all possible directions and want approximately $\mathcal{O}(10)$ points per direction, then we require $\mathcal{O}(10^5)$ points. Given that the true intrinsic dimensionality of the manifold is less than or equal to 5, this value serves as a conservative upper bound for the number of points needed to adequately sample the manifold.

@@

		Through trial and error, we observed that periodically reinitializing the optimizer significantly improved the convergence rate. This technique bears similarity to the concept of warm restarts introduced in \cite{loshchilov2017sgdrstochasticgradientdescent}. However, we found that simply scheduling the learning rate without resetting the optimizer yielded slower convergence. We interpret this as follows: when the optimizer is reinitialized, it effectively "forgets" its past gradient history. As a result, the actual learning rate used corresponds more closely to the specified value, rather than being internally adjusted based on accumulated past gradients. This effect seems to contribute to faster convergence in our case.  The evolution of the loss function, along with the learning rate schedule, is shown in Fig.~\eqref{fig:loss_grad_des}. As can be seen in this figure, the convergence rate improves significantly each time the optimizer is reinitialized. We also observe that, following the last few reinitializations, the loss exhibits a small bump immediately after the restart. We interpret this behavior as a consequence of the increased learning rate: some points that previously had low loss values may momentarily worsen before benefiting from faster convergence. This effect is likely due to points initially located in regions with weak attractive basins being pushed toward areas where the potential gradient is steeper, thus accelerating their convergence. Although this strategy introduces the risk of desampling certain regions in favor of others, they correspond to low convergence regions. As we are anyway not interested in those places, but rather real flat directions, this is not a problem for us.

@@

		As a first visualization, we present a triangular plot of the data in Fig. \ref{triangular_plot_1_2_4_8_10}. This figure shows all possible 2D projections of the data, along with the 1D histograms of each coordinate after gradient descent.
		\begin{figure}[h!]
			\centering
			\includegraphics[scale = 0.3]{Figures/triangular_plot_1_2_4_8_10.png} %names need to be changed.
			\caption{Triangular plot showing $2d$ projections and $1d$ histograms of the data after the gradient descent}
			\label{triangular_plot_1_2_4_8_10}
		\end{figure}
		We can make a few comments on those plots. First, we observe non-trivial correlations in the data. Some of these may result from larger basins of attraction, such as in the $x_1/x_2$ plot. However, the structures seen in the $x_4/\tilde{x}_8$ or $x_4/x_{10}$ plots likely reflect genuine features of the manifold, and will show latter that this is indeed the case. We also note that all directions appear to be well populated. 
		%Furthermore, we can notice that a large majority of the points converge to the origin $\vec{0}$. This is not surprising, as this just mean that the round sphere solution has a larger basin of attraction, which sounds reasonable. We do not however attempt to shift the points along flat directions to depopulate the origin or repopulate other regions.
		%\paragraph{Local PCA: graphs of number of patches with given local dimension, and with varying size of local patch, discuss optimal patch size (we want at least few points in each direction)}
		\subsubsection{Determining the dimension}
		Once the gradient descent has been completed and the flat direction sampled, the next step is to identify the structure of the underlying manifold. Our goal is to eventually obtain an analytical expression, not just a numerical description. Before applying symbolic regression to search for such an expression, we can first perform some exploratory analyses to better understand the data. Specifically, we aim to determine the dimensionality of the manifold and wether it consists of a single connected component or multiple disjoint components (e.g., two intersecting hyperplanes).
		To this end, we apply a local Principal Component Analysis (PCA). For each point, we identify its $k$ nearest neighbours and perform a PCA on that local neighbourhood. This procedure allows us to determine how many principal directions are needed to explain a given proportion $\epsilon$ of the data variance. In other words, it provides an estimate of the local dimensionality around each point.
		We perform this analysis for several values of $k$, namely $k \in \{5, 10, 20, 50, 100\}$, and we fix $\epsilon = 0.99$. The results are presented in Fig. \ref{local_pca}.

@@

		We observe that for every choice of $k$, there is a prominent peak at $d = 3$, suggesting that the underlying manifold is three-dimensional. For $k = 5$, a noticeable fraction of points are assigned a dimensionality of 2. This can be attributed to the fact that if the true dimension is 3, then selecting only 5 neighbours may not sufficiently populate all three directions, leading the algorithm to underestimate the dimensionality for a fraction of the points. 
		Additionally, for $k \geq 20$, we observe an increasing number of points being assigned dimensionalities of 4 or 5. This behavior can be explained by the loss of locality when the number of neighbours becomes too large: increasing $k$ results in a coarser approximation, and the algorithm may then incorporate points that are no longer truly local. This artificial enlargement of the neighbourhood can cause the estimated local dimensionality to rise.
		We thus conclude that the manifold under investigation has an intrinsic dimension of 3.

@@

		One possible scenario is that our data actually consists of several distinct three-dimensional manifolds, and the points previously identified with dimension $4$ may lie at the intersections of these manifolds. To illustrate, consider the intersection of two lines: at the intersection point, the local dimensionality estimated by the previous PCA algorithm would be 2. To rule out this possibility, we apply a clustering algorithm based on local density, where we only select the points of dimension 3. This, way, we remove the possible intersection points with dimension 4 or 5. For the purpose of the clustering, we use the HDBSCAN algorithm\footnote{Details of the algorithm can be found in the original HDBSCAN paper or documentation. Briefly, it is a density-based clustering method that extends DBSCAN by converting it into a hierarchical clustering algorithm and then extracting a flat clustering based on the stability of clusters.}.
		We set the minimum cluster size to 10 points. Out of the 98,606 3-dimensional points, the algorithm identifies three clusters: one of size 10, one of size 91,747, and one of size 6,849. 
		The smallest cluster (10 points) is likely an artefact of local fluctuations in the data density and is not interpreted as physically meaningful. The last group, comprising 6,849 points, consists of points that the algorithm could not assign to any cluster. We interpret these as outliers or edge points rather than representatives of a separate manifold.
		What the algorithm does indicate, however, is that over 90\% of the data belongs to a single dominant cluster. In Fig.~\ref{3dplots}, we show 3D projections of the data to visualize the clustering. 
		\begin{figure}[h!]
			\centering
			\hspace{-2cm}
			\includegraphics[scale=0.5]{Figures/3d_plot_x4_x8_x10.png}
			\includegraphics[scale=0.5]{Figures/3d_plot_x2_x4_x8.png}
			\caption{3D plots of the data in selected coordinates. Left: $(x_4, x_8, x_{10})$. Right: $(x_2, x_4, x_8)$.}
			\label{3dplots}
		\end{figure}
		In these scatter plots, turquoise points belong to the main cluster, while purple points are those that the algorithm failed to assign. In one of the plots, a few yellow points can be seen; these correspond to the smallest cluster of 10 points. From visual inspection, it appears that the unassigned (purple) points lie mostly on the boundary of the sampled region. We therefore interpret their unassigned status not as evidence of belonging to another manifold, but rather as a result of insufficient local density near the edges of the dataset.

@@

		After performing both PCA and clustering, we thus confirm that the data obtained after gradient descent samples a three-dimensional manifold.

@@

		\subsubsection{Annealed Sequential Monte Carlo principle}
		To address these challenges, we employ an Annealed Importance Sampling (AIS) combined with Sequential Monte Carlo (SMC) methods. AIS constructs a sequence of intermediate distributions that smoothly transition from an initial, tractable distribution (e.g., a prior over symbolic expressions) to the complex target posterior distribution. This annealing process is guided by a temperature-like parameter that gradually emphasizes the data likelihood, allowing for more efficient exploration of the probability landscape.

@@

		Let's us now focus on the forward $q(z_n^k | z_{n-1}^k)$ and backward propagation kernel $q(z_{n-1}^k | z_n^k)$ for now. The forward propagation kernel defines how we generate the state at time $n$ given the state at time $n-1$. It gives the probability to transition from $z_{n-1}$ to $z_n$. The backward kernel represents a hypothetical probability of transitioning back from state $z_n$ to state $z_{n-1}$. The purpose of $q$ is to give a proposition for $z_n$ given $z_{n-1}$. For the implementation of the two, we decide to implement an AIS-style MCMC code : we make some move in the space of polynomials, and then accept or reject those new polynomials based on an acceptance rate. So we first need to choose what are the available moves in the space of polynomials. Here are the choices we have made for our symbolic regression task : 
		\begin{itemize}
			\item Coefficient pertubation. Given a polynomial $z_{n-1}$, we choose randomly one of its coefficient and modify it by a random noise from the guassian distribution $\mathcal{N}(0,\sigma^2)$. So for eg : $2 x_1 + 3 x_2^2 \rightarrow  2.1 x_1 + 3 x_2^2$.
			\item Variable multiplication. Given a polynomial $z_{n-1}$, we choose randomly one of his monomial, and multiply it by one of the available variable. So for eg : $2 x_1 + 3 x_2^2 \rightarrow  2 x_1 x_2 + 3 x_2^2$.
			\item Variable division. Given a polynomial $z_{n-1}$, we choose randomly one of his monomial, and divide it by one of its variable. So for eg : $2 x_1 + 3 x_2^2 \rightarrow  2 x_1 x_2 + 3 x_2^2$. 
		\end{itemize}
		This implementation enables a direct computation of both the forward and backward propagation kernels. In particular, for the case of coefficient perturbations, the transition is symmetric. That is,
		\begin{equation}
			q(z_n^k \mid z_{n-1}^k) = q(z_{n-1}^k \mid z_n^k),
		\end{equation}
		which significantly simplifies the computation of the incremental importance weight.

>>>>

{\Large Machine learning the Conformal Manifold of Holographic CFT2’s}

@@

From this perspective, the AdS/CFT correspondence relates conformal manifolds on the boundary theory to flat directions in the scalar potential of the dual supergravity theory. Along these directions, the scalar field configurations vary continuously while the cosmological constant remains fixed. In general, supersymmetry is believed to be necessary for the existence of holographic conformal manifolds, as non-supersymmetric AdS solutions are typically expected to be unstable \cite{Ooguri:2016pdq,Palti:2019pca}. However, recent investigations \cite{Giambrone:2021wsm} have identified AdS$_4$ configurations that appear to evade this requirement, with no evidence of standard decay channels—neither perturbative nor non-perturbative—being present. In the context of AdS$_3$/CFT$_2$, the situation may be even richer. A well-known counterexample has existed for some time \cite{Aharony:2001dp,Dong:2014tsa}, which is well understood from both the field theory and gravitational perspectives. More generally, within the supergravity approximation, continuous deformations correspond to classically marginal operators. These operators become exactly marginal only if their marginality persists at finite values of both the rank $N$ and the CFT coupling. Recent work \cite{Eloy:2024lwn} has demonstrated the existence of families of marginal deformations that are perturbatively stable, using the framework of Exceptional Field Theory (ExFT).

In this work, we propose a novel approach, where we study directly the scalar potential of supergravity. As already mentionned, the classical marginal deformations of CFT's in the large $N$ limit, are in correspondance with the flat directions in the potential, i.e. when there is a continuous set a scalar fields such that $\nabla V = 0$. However, the explicit characterization of these flat directions presents formidable technical challenges. Supergravity scalar potentials, even in truncated models, typically involve dozens of scalar fields with intricate non-linear interactions. The resulting expressions for critical points—where all first derivatives vanish—quickly become too complex for traditional symbolic manipulation, rendering analytical approaches computationally intractable.

The emergence of machine learning techniques in theoretical physics opens new avenues for addressing such complex problems. Instead of solving the full symbolic system analytically from the outset, one can employ numerical methods to sample the solution space and subsequently apply symbolic regression techniques to extract analytical patterns from the data. This hybrid approach has the potential to bypass the computational bottlenecks inherent in purely symbolic methods, while still having the potential to uncover exact analytical expressions.

Machine learning strategies have previously been applied to identify new isolated vacua in $\mathrm{SO}(8)$ supergravity \cite{Comsa:2019rcz,Berman:2022jqn}. More broadly, there has been increasing interest in applying machine learning and numerical techniques across various domains of high-energy physics. This includes, for instance, the characterization of Calabi–Yau metrics and hypersurfaces \cite{Ashmore:2019wzb,Berman:2021mcw,Larfors:2022nep,Berglund:2022gvm,Jejjala:2020wcc,Douglas:2006rr,Larfors:2021pbb,He:2018jtw}, as well as broader efforts to explore the string theory landscape using machine learning techniques \cite{He:2017aed,Carifio:2017bov,Ruehle:2017mzq}. Additional applications include studies in CFT \cite{Chen:2020dxg}, investigations of the supergravity landscape \cite{Brady:2025zzi,Krishnan:2020sfg}, and explorations of the AdS/CFT correspondence \cite{Hashimoto:2018ftp}. More generally, machine learning has found utility in the study of string theory, geometry, and fundamental physics \cite{Ruehle:2020jrk,He:2023csq,Bao:2021auj}.

In this work, we demonstrate the viability of the aformentionned machine learning approach by applying it to a five-scalar subsector of a 13 scalar consistent truncation of six-dimensional non-chiral $\mathcal{N} = (1,1)$ supergravity on $AdS_3 \times S^3$, or to type IIB supergravity on $AdS_3 \times S^3 \times T^4$. Our methodology combines gradient descent sampling of the flat direction manifold with some symbolic regression technique. There exist a large litterature on symbolic regression, using methods from genetic programming \cite{koza1994genetic} such as \cite{virgolin2021improving,randall2022bingo,burlacu2019parsimony}, to deep machine learning \cite{petersen2019deep,kamienny2022end}, generative modelling \cite{valipour2021symbolicgpt}, diffusion models \cite{bastiani2025diffusion}, and equation learner with the nodes being symbolic operations \cite{2018arXiv180607259S}. Another state-of-the art algorithm is the AIFeynman methods \cite{Udrescu:2019mnk}. They use neural network to identify structure in the dataset (such as translational symmetries, multiplicative separability, compositionality, \dots) to recursively define simpler problems on which they can fit the solutions with polynomials.  However, in the spirit of trying to build a generalisable method, we develop here our own algorithm, which is based on symbolic regression using an Annealing Importance Sampling method (AIS) \cite{neal1998annealedimportancesampling}. In this paper, we will study a 5 parameter truncation, which is itself coming from a 13 parameter consistent truncation of $\mathcal{N} = (1,1)$ 6-dimensional supergravity.

The study of this 5 dimensional parameter potentiel is separated into several parts. We first use the gradient descent to efficiently sample the underlying conformal manifold. Combined with numerical analysis, such as principal component analysis and clustering, we can identify the existence of a 3-dimensional continuous family. We then use an AIS technique combined with a sequential Monte Carlo approach (SMC) \cite{del2006sequential} in order to do symbolic regression. As we will indeed demonstrate in the paper, there exist polynomial constraints on the 5 parameters, viewed as embedding coordinates, that projects them on the 3 dimensional manifolds. We manage to identify 8 of those constraints on the data, not all independent, which once solved gives that two of the embedding coordinates can be expressed in terms of the others. This provides an explicit 3 dimensional parameters family

The paper is organized as follows. Section 2 establishes the supergravity setup, presenting the scalar potential in its full complexity and motivating the truncation to five fields. Section 3 details our numerical methodology, covering both the gradient descent sampling procedure and the annealed importance sampling approach to symbolic regression. Finally, Section 4 concludes with prospects for extending this approach to higher-dimensional cases and its broader implications for systematic studies of conformal manifolds in holographic theories.

@@

The Chern-Simons contribution is given by
%
\begin{equation}
	{\cal L}_{\rm CS}= -\varepsilon^{\,\mu\nu\rho}\,\Theta_{\bar M\bar N|\bar P\bar Q}\,A_{\mu}{}^{\bar M\bar N}\left(\partial_{\nu}\,A_{\rho}{}^{\bar P\bar Q}  + \frac{1}{3}\, \Theta_{\bar R\bar S|\bar U\bar V}\,f^{\bar P\bar Q,\bar R\bar S}{}_{\bar X\bar Y}\, A_{\nu}{}^{\bar U\bar V} A_{\rho}{}^{\bar X\bar Y} \right)\,,
\end{equation}
%
where $f^{\bar M\bar N,\bar P\bar Q}{}_{\bar K\bar L} = 4\,\delta_{[\bar K}{}^{[\bar M}\eta^{\bar N][\bar P}\delta_{\bar L]}{}^{\bar Q]}$ are the structure constants of the SO(8,4) Lie algebra, and $\varepsilon^{\mu\nu\rho}$ is the three-dimensional Levi-Civita symbol.

@@

A first numerical exploration suggests that attention can be focused on the five scalars $x_{1}, x_{2}, x_{4}, \tilde{x}_{8}$, and $x_{10}$. This subset, selected at random from the original set of thirteen scalar fields, serves to render the problem more tractable at the outset. It is important to emphasize, however, that this subset does not constitute a consistent truncation of the full theory. Consequently, the truncated theory involving only five scalar fields may not constitute a formally consistent truncation of the full 13-scalar theory, as higher-order interactions could, in principle, source the remaining scalar modes. As a result, the interpretation of the five-scalar theory as a consistent subsector of the original theory may no longer hold. Nonetheless, the five-scalar model remains a self-consistent theory in its own right and can be used independently. This distinction does not affect our numerical approach.

@@

		To do so, we perform a basic gradient descent. We begin by randomly and uniformly initializing points within a hypercube of range $[-2,2]$. This choice is important to ensure that points are not restricted to the inner range $[-1,1]$. This will be important later, when performing the symbolic regression \footnote{The rationale is that symbolic regression (SR) evaluates the data points ${\vec{x}^i}$ using a candidate polynomial function $p$. The loss function—used later to assess the quality of a given polynomial—is defined as $\sum_i p(\vec{x}^i)$. When all data points lie within the interval $[-1, 1]$, the SR algorithm tends to favor high-degree polynomials, as these often yield smaller values over this range, thereby artificially lowering the loss. Introducing data points with absolute values greater than 1 mitigates this bias by discouraging the selection of overly complex polynomials that exploit this effect.}. For this example, we choose to generate $10^5$ points. This number is motivated by the dimensionality of our problem. If we aim to populate all possible directions and want approximately $\mathcal{O}(10)$ points per direction, then we require $\mathcal{O}(10^5)$ points. Given that the true intrinsic dimensionality of the manifold is less than or equal to 5, this value serves as a conservative upper bound for the number of points needed to adequately sample the manifold.

@@

		Through trial and error, we observed that periodically reinitializing the optimizer significantly improved the convergence rate. This technique bears similarity to the concept of warm restarts introduced in \cite{loshchilov2017sgdrstochasticgradientdescent}. However, we found that simply scheduling the learning rate without resetting the optimizer yielded slower convergence. We interpret this as follows: when the optimizer is reinitialized, it effectively "forgets" its past gradient history. As a result, the actual learning rate used corresponds more closely to the specified value, rather than being internally adjusted based on accumulated past gradients. This effect seems to contribute to faster convergence in our case.  The evolution of the loss function, along with the learning rate schedule, is shown in Fig.~\eqref{fig:loss_grad_des}. As can be seen in this figure, the convergence rate improves significantly each time the optimizer is reinitialized. We also observe that, following the last few reinitializations, the loss exhibits a small bump immediately after the restart. We interpret this behavior as a consequence of the increased learning rate: some points that previously had low loss values may momentarily worsen before benefiting from faster convergence. This effect is likely due to points initially located in regions with weak attractive basins being pushed toward areas where the potential gradient is steeper, thus accelerating their convergence. This strategy introduces the risk of desampling certain regions in favor of others, introducing the risk to hide some flat directions in the potential. However, we believe that with enough points, it is very unlikely for a flat direction to be completely hidden from us and desampled.

@@

		As a first visualization, we present a triangular plot of the data in Fig. \ref{triangular_plot_1_2_4_8_10}. This figure shows all possible 2D projections of the data, along with the 1D histograms of each coordinate after gradient descent.
		\begin{figure}[h!]
			\centering
			\includegraphics[scale = 0.3]{Figures/triangular_plot_1_2_4_8_10.png} %names need to be changed.
			\caption{Triangular plot showing $2d$ projections and $1d$ histograms of the data after the gradient descent}
			\label{triangular_plot_1_2_4_8_10}
		\end{figure}
		We can make a few comments on those plots. First, we observe non-trivial correlations in the data. Some of these may result from larger basins of attraction, such as in the $x_1/x_2$ plot. However, the structures seen in the $x_4/\tilde{x}_8$ or $x_4/x_{10}$ plots likely reflect genuine features of the manifold, and will show latter that this is indeed the case. We also note that all directions appear to be well populated. 
		%Furthermore, we can notice that a large majority of the points converge to the origin $\vec{0}$. This is not surprising, as this just mean that the round sphere solution has a larger basin of attraction, which sounds reasonable. We do not however attempt to shift the points along flat directions to depopulate the origin or repopulate other regions.
		%\paragraph{Local PCA: graphs of number of patches with given local dimension, and with varying size of local patch, discuss optimal patch size (we want at least few points in each direction)}
		\subsubsection{Determining the dimension}
		Once the gradient descent has been completed and the flat directions sampled, the next step is to identify the structure of the underlying manifold. Our goal is to eventually obtain an analytical expression, not just a numerical description. Before applying symbolic regression to search for such an expression, we can first perform some exploratory analyses to better understand the data. Specifically, we aim to determine the dimensionality of the manifold and wether it consists of a single connected component or multiple disjoint components (e.g., two intersecting hyperplanes).
		To this end, we apply a local Principal Component Analysis (PCA). For each point, we identify its $k$ nearest neighbours and perform a PCA on that local neighbourhood. This procedure allows us to determine how many principal directions are needed to explain a given proportion $\epsilon$ of the data variance. In other words, it provides an estimate of the local dimensionality around each point.
		We perform this analysis for several values of $k$, namely $k \in \{5, 10, 20, 50, 100\}$, and we fix $\epsilon = 0.99$. The results are presented in Fig. \ref{local_pca}.

@@

		We observe that for every choice of $k$, there is a prominent peak at $d = 3$, suggesting that the underlying manifold is three-dimensional. For $k = 5$, a noticeable fraction of points are assigned a dimensionality of 2. This can be attributed to the fact that if the true dimension is 3, then selecting only 5 neighbours may not sufficiently populate all three directions, leading the algorithm to underestimate the dimensionality for a fraction of the points. 
		Additionally, for $k \geq 20$, we observe an increasing number of points being assigned dimensionalities of 4 or 5. This behavior can be explained by the loss of locality when the number of neighbours becomes too large: increasing $k$ results in a coarser approximation, and the algorithm may then incorporate points that are no longer truly local. This artificial enlargement of the neighbourhood can cause the estimated local dimensionality to rise.
		We thus assume that the manifold under investigation has an intrinsic dimension of 3.

@@

		One possible scenario is that our data actually consists of several distinct three-dimensional manifolds, and the points previously identified with dimension $4$ may lie at the intersections of these manifolds. To illustrate, consider the intersection of two lines: at the intersection point, the local dimensionality estimated by the previous PCA algorithm would be 2. To rule out this possibility, we apply a clustering algorithm based on local density, where we only select the points of dimension 3. This, way, we remove the possible intersection points with dimension 4 or 5. For the purpose of the clustering, we use the HDBSCAN algorithm\footnote{Details of the algorithm can be found in the original HDBSCAN paper or documentation \cite{10.1007/978-3-642-37456-2_14}. Briefly, it is a density-based clustering method that extends DBSCAN by converting it into a hierarchical clustering algorithm and then extracting a flat clustering based on the stability of clusters.}.
		We set the minimum cluster size to 10 points. Out of the 98,606 3-dimensional points, the algorithm identifies three clusters: one of size 10, one of size 91,747, and one of size 6,849. 
		The smallest cluster (10 points) is likely an artefact of local fluctuations in the data density and is not interpreted as physically meaningful. The last group, comprising 6,849 points, consists of points that the algorithm could not assign to any cluster. We interpret these as outliers or edge points rather than representatives of a separate manifold.
		What the algorithm does indicate, however, is that over 90\% of the data belongs to a single dominant cluster. In Fig.~\ref{3dplots}, we show 3D projections of the data to visualize the clustering. 
		\begin{figure}[h!]
			\centering
			\hspace{-2cm}
			\includegraphics[scale=0.5]{Figures/3d_plot_x4_x8_x10.png}
			\includegraphics[scale=0.5]{Figures/3d_plot_x2_x4_x8.png}
			\caption{3D plots of the data in selected coordinates. Left: $(x_4, x_8, x_{10})$. Right: $(x_2, x_4, x_8)$.}
			\label{3dplots}
		\end{figure}
		In these scatter plots, turquoise points belong to the main cluster, while purple points are those that the algorithm failed to assign. In one of the plots, a few yellow points can be seen; these correspond to the smallest cluster of 10 points. From visual inspection, it appears that the unassigned (purple) points lie mostly on the boundary of the sampled region. We therefore interpret their unassigned status not as evidence of belonging to another manifold, but rather as a result of insufficient local density near the edges of the dataset.

@@

		After performing both PCA and clustering, we thus find that the data obtained after gradient descent samples a three-dimensional manifold.

@@

		\subsubsection{Annealed Sequential Monte Carlo principle}

		\begin{figure}[h!]
			\centering
			\begin{minipage}{15cm}
				\begin{tcolorbox}[
					colback=white,
					colframe=myblue,
					title=\textbf{ASMC Algorithm},
					fonttitle=\sffamily\large,
					halign title=flush center,
					boxrule=2pt,
					left=6pt,
					right=6pt,
					top=8pt,
					bottom=8pt
					]
					\begin{tikzpicture}[node distance=1.6cm and 0.8cm, scale=0.85, transform shape]

						\node (ess_check) [decision, below of=weight_update] {Effective Sample Size $<$ Threshold ?};

						\node (mutation) [process, below of=ess_check, yshift=-0.4cm] {Mutate particles with MCMC kernel};

						\node (output) [process, below of=exploit] {Return particles};

						% Boucle de retour améliorée
						\draw [arrow] (end_iter) -| node[pos=0.25, below, arrow_label] {No} ++(-4.5,0) |- (compute_beta);

		To address these challenges, we employ an Annealed Importance Sampling (AIS) combined with Sequential Monte Carlo (SMC) methods. AIS constructs a sequence of intermediate distributions that smoothly transition from an initial, tractable distribution (e.g., a prior over symbolic expressions) to the complex target posterior distribution. This annealing process is guided by a temperature-like parameter that gradually emphasizes the data likelihood, allowing for more efficient exploration of the probability landscape.

@@

		Let's us now focus on the forward $q(z_n^k | z_{n-1}^k)$ and backward propagation kernel $q(z_{n-1}^k | z_n^k)$ for now. The forward propagation kernel defines how we generate the state at time $n$ given the state at time $n-1$. It gives the probability to transition from $z_{n-1}$ to $z_n$. The backward kernel represents a hypothetical probability of transitioning back from state $z_n$ to state $z_{n-1}$. The purpose of $q$ is to give a proposition for $z_n$ given $z_{n-1}$. For the implementation of the two, we decide to implement an AIS-style MCMC code : we make some move in the space of polynomials, and then accept or reject those new polynomials based on an acceptance rate. So we first need to choose what are the available moves in the space of polynomials. Here are the choices we have made for our symbolic regression task : 
		\begin{itemize}
			\item Coefficient pertubation. Given a polynomial $z_{n-1}$, we choose randomly one of its coefficient and modify it by a random noise from the guassian distribution $\mathcal{N}(0,\sigma^2)$. So for eg : $2 x_1 + 3 x_2^2 \rightarrow  2.1 x_1 + 3 x_2^2$.
			\item Variable multiplication. Given a polynomial $z_{n-1}$, we choose randomly one of his monomial, and multiply it by one of the available variable. So for eg : $2 x_1 + 3 x_2^2 \rightarrow  2 x_1 x_2 + 3 x_2^2$.
			\item Variable division. Given a polynomial $z_{n-1}$, we choose randomly one of his monomial, and divide it by one of its variable. So for eg : $2 x_1 x_2 + 3 x_2^2 \rightarrow  2 x_1 + 3 x_2^2$. 
		\end{itemize}
		This implementation enables a direct computation of both the forward and backward propagation kernels. In particular, for the case of coefficient perturbations, the transition is symmetric. That is,
		\begin{equation}
			q(z_n^k \mid z_{n-1}^k) = q(z_{n-1}^k \mid z_n^k),
		\end{equation}
		which significantly simplifies the computation of the incremental importance weight.
